---
title: "Lecture 5 - Data Import"
output: html_document
---

# Random Notes
*R can hold integer, character, vector and factor variables
*Data frames are actually lists.
    tibble = a version of a dataframe that is easier to print  
*In general, try to get the most raw form of data so that you have complete control over the cleaning process 
which is easy to fix/double check later on

# Importing Files/Directories

When importing files from your computer, you call upon a file's directory/storage pathway, which can be either: 

*Absolute: a file or folder’s “full address” on your computer
    e.g. "/Users/jeffgoldsmith/Dropbox/Work/Teaching/P8105/Fall_2018/p8105.github.io/data/FAS_litters.csv"
  
*Relative: directions to a file or folder from your current working directory. to first find your working directory where this project is stored, type 'getwd()' and represent this with a period . before adding /... to specify exactly where your file is located within the working directory  
    e.g. "./data/FAS_litters.csv"

Note: it's better to use a relative directory bc its easier for someone else to access it without having the 
exact same pathway as you. but if you do send the rproject file to someone, you should explain how it's saved 

Shorthand notes for relative directories: 
~ = home directory
. = current working directory
.. = one directory up from current working directory
../.. two directories up from current working directory 

# Importing Data Tables

we're using csv comma delimited files (csv)
```{r}
library(tidyverse)
getwd()
litters_data = read_csv(file = "./sample data/FAS_litters.csv")
litters_data = janitor::clean_names(litters_data) # to clean up var names 
```


# Looking at Data
```{r}
names(litters_data) #to see all variable names

# use any of the following commands to view the dataset in the environment window on the right: 
str
head (litters_data, 5) # to see first five rows of data  
View
tail

skimr::skim(litters_data) # to generate summary stats e.g. # of obs, # of vars
```

# How to read a data file

```{r}
litters_data = read_csv(file = "./sample data/FAS_litters.csv",
  skip = 10, col_names = FALSE) #Skip indicates the # of rows to skip before reading data. col_names = FALSE or TRUE indicates if you want to keep (true) or reject (false) the existing column/var names in which case they'll automatically show up as X1 and you can then change it to a new column name. Usually you'd use TRUE if the first row contains variable names 

head(litters_data) #to double check how the first few lines look
```

# Parsing Columns 

if you parse columns, make sure to be very specific about labeling what kind of variable each is.. e.g. character, integer, etc. Although not necessary because 'col_names' usually does a good job automatically guessing the type of data stored in each column 
```{r}
litters_data = read_csv(file = "./sample data/FAS_litters.csv",
  col_types = cols(
    Group = col_character(),
    `Litter Number` = col_character(),
    `GD0 weight` = col_double(), #double = numeric type of data that CANNOT contain decimals
    `GD12 weight` = col_double(),
    `GD of Birth` = col_integer(), #integer = numeric type of data CAN contain decimals 
    `Pups born alive` = col_integer(),
    `Pups dead @ birth` = col_integer(),
    `Pups survive` = col_integer()
  )
)
tail(litters_data)

litters_data = read_csv(file = "./sample data/FAS_litters.csv",
  col_types = "ccddiiii"
)
litters_data
```

# Other file formats

To import non-csv, plain text files like excel files use 'read_table' instead of 'read_csv'
```{r}
library(readxl) # to load excel package
mlb11_data = read_excel("./sample data/mlb11.xlsx", n_max = 20)
head(mlb11_data, 5)
```

To import files from SAS, STATA, and SPSS. 
```{r}
library(haven) #to load haven package
pulse_data = read_sas("./sample data/public_pulse_data.sas7bdat")
head(pulse_data, 5)
```

# Comparison with Base R
'read_csv' produces tibbles that are very similar to the base R data frames produced by 'read.csv'. However, tibbles have some features that can help prevent mistakes and unwanted behavior.

# Learning Assessment
```{r}
pups_base = read.csv("./sample data/FAS_pups.csv")
pups_readr = read_csv("./sample data/FAS_pups.csv")

View(pups_base)
View(pups_readr) #decimals show up instead of spaces between words

pups_base #The litter number column is assigned as a factor type 
pups_readr #The litter number column is assigned as a character type. Also, decimals show up instead of spaces between words.

pups_base$S
pups_readr$S #this command doesn't work... doesn't recognize 'S'
```

# Export Data
use 'write_' functions in readr 



